---
title: Supply Chain Delivery Analysis
author: Casey Ortiz
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    theme: flatly
    highlight: tango
---

<!-- ✅ Custom banner -->
<div style="background-color:#2C3E50; color:white; padding:20px; border-radius:6px; margin:30px 0 20px 0; text-align:center;">
  <h2 style="margin:0 0 8px 0;">Supply Chain Delivery Analysis</h2>
  <p style="margin:0;">Case Study by Casey Ortiz — <strong>`r Sys.Date()`</strong></p>
</div>

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Executive Summary

This case study explores on-time delivery performance within a supply chain logistics dataset. Using a combination of data cleaning, exploratory analysis, and machine learning modeling in R and Python, we set out to identify key drivers behind late shipments and uncover operational insights to reduce delivery risk.

We analyzed over 10,000 records across multiple shipment types, product weights, and delivery channels. Our models achieved up to **68% accuracy**, but the most valuable discovery came through segmentation:  

* **Products under $175 and less than 4 lbs** had a **96.8% on-time delivery rate**, while  
* **Products between 4.5–8 lbs** showed a **0% late delivery rate**.

Further analysis of transportation modes and warehouse types revealed clear operational patterns within these high-performing segments.

As a result, we recommend optimizing fulfillment routing and inventory placement strategies for light, low-cost items — and prioritizing warehouse and mode consistency to further reduce delivery delays.

```{r executive-visual, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, dpi=300}

# 📦 Load packages
library(tidyverse)

# 🔁 Load the cleaned data
data <- read_csv("train.csv")

# 🔁 Recode delivery status
data <- data %>%
  mutate(on_time_delivery = case_when(
    on_time_delivery %in% c(1, "1") ~ "On Time",
    on_time_delivery %in% c(0, "0") ~ "Late",
    TRUE ~ NA_character_
  )) %>%
  mutate(on_time_delivery = factor(on_time_delivery, levels = c("Late", "On Time")))

# 🎯 Plot with styled legend inside top-left
ggplot(data, aes(x = product_cost, y = weight_lbs, color = on_time_delivery)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = c("Late" = "#F8766D", "On Time" = "#00BA38")) +
  labs(
    title = "📊 Product Cost vs Weight by Delivery Outcome",
    x = "Product Cost ($)",
    y = "Weight (lbs)",
    color = "Delivery Status"
  ) +
  annotate("text", x = 45, y = 7.3, label = "The Big Mystery",
           color = "black", size = 8, hjust = 0, fontface = "bold") +
  annotate("text", x = 3.6, y = 3.3, label = "Near Perfect: <$175 & <4.5lbs.",
           color = "black", size = 4.5, hjust = 0, fontface = "bold") +
  annotate("rect", xmin = 0, xmax = 175, ymin = 2, ymax = 4.5,
           alpha = 0.1, fill = "#2CA02C") +
  annotate("text", x = 41, y = 5.8, label = "Perfect Delivery Rate: 4.5–8 lbs",
           color = "black", size = 4.5, hjust = 0, fontface = "bold") +
  annotate("rect", xmin = 0, xmax = 285, ymin = 4.5, ymax = 8.8,
           alpha = 0.08, fill = "#1f77b4") +
  coord_cartesian(xlim = c(0, 315)) +
  theme_minimal() +
  theme(
    legend.position = c(0.05, 0.95),  # Top-left inside
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = alpha("white", 0.8), color = "gray70"),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14)
  )
```

# Case Study: Predicting Late Deliveries in Supply Chain Logistics

## 1. Problem Statement

Late deliveries reduce customer satisfaction and increase operational
costs. The goal was to analyze supply chain performance data and build a
predictive model to identify which shipments are at risk of arriving
late.

------------------------------------------------------------------------

## 2. Dataset Overview

-   Dataset: Kaggle – E-Commerce Shipping Data\
-   10,999 rows representing individual shipments\
-   Key features: delivery_type, warehouse_type, delivery_priority, weight_lbs\
-   Target variable: `on_time_delivery` (1 = on time, 0 = late)

------------------------------------------------------------------------

## 3. Data Cleaning and Preparation

-   Standardized all column names (snake_case)
-   Converted weight from grams to pounds (rounded to 2 decimals)
-   Label-encoded categorical variables
-   Binned `product_cost` and `customer_calls` for pattern discovery
-   Dropped identifier column
-   Exported cleaned dataset for modeling

```{r Data Preparation and Cleaning}

# Load libraries
library(tidyverse)
library(readr)
library(lubridate)
library(ggplot2)
library(janitor)

# Load dataset 
data <- read_csv("train_raw.csv")
# View(data)  # Disable this line when knitting


summary(data)
glimpse(data)

# On-time delivery summary
data %>%
  group_by(Reached.on.Time_Y.N) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

# Rename columns for clarity
data <- data %>%
  rename(
    id = ID,
    warehouse_type = Warehouse_block,
    delivery_type = Mode_of_Shipment,
    customer_calls = Customer_care_calls,
    customer_review = Customer_rating,
    product_cost = Cost_of_the_Product,
    prior_purchases = Prior_purchases,
    delivery_priority = Product_importance,
    gender = Gender,
    discount_offered = Discount_offered,
    weight_lbs = Weight_in_gms,
    on_time_delivery = Reached.on.Time_Y.N
  )

# Convert grams to lbs and round 2 decimal places
data <- data %>%
  mutate(weight_lbs = round(weight_lbs / 453.592, 2))

# Set column types
data <- data %>%
  mutate(
    warehouse_type = as.factor(warehouse_type),
    delivery_type = as.factor(delivery_type),
    gender = as.factor(gender),
    delivery_priority = factor(delivery_priority, levels = c("low", "medium", "high"), ordered = TRUE),
    customer_review = as.integer(customer_review),
    customer_calls = as.integer(customer_calls),
    prior_purchases = as.integer(prior_purchases),
    discount_offered = as.integer(discount_offered),
    on_time_delivery = as.factor(on_time_delivery)
  )

# Save cleaned data to CSV
write_csv(data, "train.csv")

# Summary of dataset
summary(data)

# Check levels of categorical variables
sapply(data[, sapply(data, is.factor)], levels)

# Drop id column before modeling
data <- data %>% select(-id)

#Check balance of target variable
table(data$on_time_delivery)
prop.table(table(data$on_time_delivery))

#Reorder Columns
data <- data %>%
  select(on_time_delivery, delivery_priority, delivery_type, everything())

# Export cleaned dataset (optional)
write_csv(data, "train_cleaned.csv")
```

------------------------------------------------------------------------

## 4. Exploratory Data Analysis (EDA)

-   On-time delivery rates were surprisingly consistent across priority
    levels, shipment modes, and warehouses
-   Some slight increases in late delivery rates were seen with higher
    product cost and increased customer calls

**Visuals Created:** 

* Product Cost vs Weight: Identified low-cost, low-weight segments with nearly perfect on-time delivery.
* Pie Chart by Delivery Type: Shipments in perfect segments are mostly concentrated in "ship" transportation mode.
* Pie Chart by Warehouse Type: High-performing deliveries are unevenly distributed across warehouse types, offering targeting opportunities.

```{r}
summary(data)
glimpse(data)

#Check late deliveries by category
data %>%
  group_by(delivery_priority, on_time_delivery) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

data %>%
  group_by(delivery_type, on_time_delivery) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

data %>%
  group_by(warehouse_type, on_time_delivery) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

data %>%
  group_by(prior_purchases, on_time_delivery) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))
```

------------------------------------------------------------------------

## 5. Preprocessing and Feature Engineering

The raw dataset required a series of preprocessing steps to prepare it for modeling. These included:

- Cleaning column names for readability
- Converting weight from grams to pounds
- Encoding categorical variables (e.g., warehouse type, shipment mode, delivery priority)
- Binning numerical variables like `product_cost` and `weight_lbs` for visualization
- Creating ordered factors for `delivery_priority` and `customer_review`

These steps were performed in R for data exploration, and replicated in Python to support modeling in the Colab notebook.

------------------------------------------------------------------------

## 6. Predictive Modeling (Python)

To complement the exploratory work in R, I built and evaluated a series of classification models in Python using Google Colab.

The workflow included:

- Preprocessing: Label encoding of categorical variables and stratified train-test split
- Evaluation: Accuracy, precision, recall, F1-score, and confusion matrix
- Modeling:  
  - **Logistic Regression**  
  - **Decision Tree Classifier**  
  - **XGBoost Classifier** (with hyperparameter tuning)

The decision tree and tuned XGBoost models achieved the best recall on late deliveries, although overall accuracy plateaued around 68%. Full code, metrics, and visualizations are available in the linked Colab notebook.


### Model Performance (Holdout Set)
| Model               | Accuracy | Precision (Late) | Recall (Late) | F1 Score (Late) |
|--------------------|----------|------------------|----------------|-----------------|
| Logistic Regression| 63.5%    | 54%              | 58%            | 56%             |
| Decision Tree      | 68.0%    | 56%              | 98%            | 71%             |
| XGBoost (Tuned)    | 67.9%    | 56%              | 98%            | 71%             |

> 🔍 XGBoost delivered the strongest precision-recall balance after hyperparameter tuning. Feature importance suggested that **product weight** and **cost** were the dominant drivers of delivery performance.

### Access Full Modeling Notebook
📎 [View the full Colab notebook](https://colab.research.google.com/drive/1FVgIY6qvmbH7W0Zrpdu1KYgv2V9KvMJ5?usp=sharing)  
*(Includes code, training pipeline, and visualizations)*

## 7. Visual Insights Summary

The following charts illustrate the key findings that support the modeling and recommendations:

- 📊 *Product Cost vs Weight*: Identified low-cost, low-weight segments with nearly perfect on-time delivery.
- 🧁 *Pie Chart by Delivery Type*: Shipments in perfect segments are mostly concentrated in [top categories].
- 🏬 *Pie Chart by Warehouse Type*: High-performing deliveries are unevenly distributed across warehouse blocks, offering targeting opportunities.


```{r Product Cost vs Weight by Delivery Outcome Scatterplot, fig.width=10, fig.height=6, dpi=300}

# 📦 Load packages
library(tidyverse)

# 🔁 Load the cleaned data
data <- read_csv("train.csv")

# 🔁 Recode delivery status
data <- data %>%
  mutate(on_time_delivery = case_when(
    on_time_delivery %in% c(1, "1") ~ "On Time",
    on_time_delivery %in% c(0, "0") ~ "Late",
    TRUE ~ NA_character_
  )) %>%
  mutate(on_time_delivery = factor(on_time_delivery, levels = c("Late", "On Time")))

# 🎯 Plot with styled legend inside top-left
ggplot(data, aes(x = product_cost, y = weight_lbs, color = on_time_delivery)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = c("Late" = "#F8766D", "On Time" = "#00BA38")) +
  labs(
    title = "📊 Product Cost vs Weight by Delivery Outcome",
    x = "Product Cost ($)",
    y = "Weight (lbs)",
    color = "Delivery Status"
  ) +
  annotate("text", x = 45, y = 7.3, label = "The Big Mystery",
           color = "black", size = 8, hjust = 0, fontface = "bold") +
  annotate("text", x = 3.6, y = 3.3, label = "Near Perfect: <$175 & <4.5lbs.",
           color = "black", size = 4.5, hjust = 0, fontface = "bold") +
  annotate("rect", xmin = 0, xmax = 175, ymin = 2, ymax = 4.5,
           alpha = 0.1, fill = "#2CA02C") +
  annotate("text", x = 41, y = 5.8, label = "Perfect Delivery Rate: 4.5–8 lbs",
           color = "black", size = 4.5, hjust = 0, fontface = "bold") +
  annotate("rect", xmin = 0, xmax = 285, ymin = 4.5, ymax = 8.8,
           alpha = 0.08, fill = "#1f77b4") +
  coord_cartesian(xlim = c(0, 315)) +
  theme_minimal() +
  theme(
    legend.position = c(0.05, 0.95),  # Top-left inside
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = alpha("white", 0.8), color = "gray70"),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14)
  )

# ✅ Save plot
ggsave("product_cost_vs_weight.png", width = 10, height = 6, dpi = 300)

```

```{r}
# Filter for nearly perfect deliveries (e.g., low cost & low weight OR ideal weight zone)
perfect_subset <- data %>%
  filter((product_cost < 175 & weight_lbs < 4) | (weight_lbs >= 4.5 & weight_lbs <= 8))

# Summarize proportions by delivery type
delivery_type_dist <- perfect_subset %>%
  group_by(delivery_type) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

# Pie chart
ggplot(delivery_type_dist, aes(x = "", y = percent, fill = delivery_type)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(percent, "%")), position = position_stack(vjust = 0.5), color = "white") +
  labs(
    title = "📦 On-Time Delivery by Delivery Type (Perfect Segments)",
    fill = "Delivery Type"
  ) +
  theme_void()

# ✅ Save plot
ggsave("on_time_delivery_by_delivery_type.png", width = 10, height = 6, dpi = 300)

#library(scales)
#ggplot_build(p)$data[[1]]$fill  # For fill colors

```

```{r}
# 🎨 Define the color palette for all 5 warehouse types
warehouse_colors <- c(
  "A" = "#F8766D",
  "B" = "#00BA38",
  "F" = "#619CFF",
  "D" = "#E68613",
  "C" = "#956CB4"
)

# 📊 Summarize proportions by warehouse type
warehouse_dist <- perfect_subset %>%
  group_by(warehouse_type) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 1))

# 🧁 Pie chart
ggplot(warehouse_dist, aes(x = "", y = percent, fill = warehouse_type)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(percent, "%")),
            position = position_stack(vjust = 0.5), color = "white") +
  scale_fill_manual(values = warehouse_colors) +
  labs(
    title = "🏬 On-Time Delivery by Warehouse Type (Perfect Segments)",
    fill = "Warehouse"
  ) +
  theme_void()

# ✅ Save plot
ggsave("on_time_delivery_by_warehouse_type.png", width = 10, height = 6, dpi = 300)
```

------------------------------------------------------------------------

## 8. Final Insights and Recommendations

Data analysis and modeling surfaced several high-value patterns:

### High-Performance Delivery Segments
- Products under **$175 and under 4 lbs** achieved a **96.8% on-time delivery rate**
- Products between **4.5–8 lbs** showed **0% late deliveries**
- These items clustered around select warehouses (notably Type F) and primarily used “Ship” as delivery mode

### Operational Recommendations
- **Replicate perfect conditions**: Investigate which SKUs fall into this group. Are there packaging, route, or facility advantages that can be scaled?
- **Audit high-performing warehouses**: Type F warehouses had a strong presence in the best-performing segments. Audit these facilities to identify operational best practices. Consider prioritizing their use for high-risk deliveries.
- **Validate delivery mode logic**: “Ship” shipments account for ~70% of the on time deliveries in our subset. Validate how modes (Road, Ship, Flight) are assigned and evaluate average transit times. If shipping mode correlates with geography or warehouse location, re-optimizing routing logic may improve delivery accuracy.

### Strategic Modeling Use
- Use model predictions to **flag high-risk deliveries** based on cost, weight, and discounts — deprioritize low-signal features like gender
- Employ business rules post-model to escalate high-priority customer orders, even if model confidence is high


------------------------------------------------------------------------

## 9. Next Steps

To operationalize these insights and extend model value:

- **Refine business context**  
  Collaborate with operations teams to validate routing logic, facility constraints, and packaging practices

- **Hyperparameter tuning**  
  Further optimize XGBoost or LightGBM using Optuna to improve recall on late deliveries

- **Enrich the dataset**  
  Add external signals like weather, holidays, region, or real-time delay feeds to boost model accuracy

- **Deploy a real-time dashboard**  
  Use Tableau, Streamlit, or Power BI to surface predicted late deliveries and monitor impact

- **Test model-driven routing decisions**  
  Run A/B tests or operational pilots to evaluate the real-world business impact of acting on predictions
  
------------------------------------------------------------------------  
  
## 10. Project Files & Portfolio Links

GitHub Repository: [github.com/yourusername/supply-chain-delivery](#)  
Google Colab Notebook: [View in Colab](https://colab.research.google.com/drive/1FVgIY6qvmbH7W0Zrpdu1KYgv2V9KvMJ5?usp=sharing)  
R Markdown Report (HTML): [View HTML](#)  
Key Visuals: [product_cost_vs_weight.png](#), [warehouse_type_pie.png](#)

---

This project demonstrates applied analytics across the full pipeline — from data cleaning and EDA in R to modeling, segmentation, and operational recommendations using Python and visualization tools. For recruiters or teams reviewing this case study, I’m happy to walk through the notebook and share additional insights.

